---
title: "Review of Seasons for Ethiopia"
author: "Pauline"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Rainy Season Review by Admin Level

This document looks at possible seasons per each admin level for Ethiopia.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(TriggerApp2024)
library(tidyverse)
library(gghdx)
gghdx()
#ldf <- load_df_forecast(dataset = "mars_eth")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
eth_files <- stringr::str_subset(list.files('../data'), pattern = "eth")
fp_eth_files <- file.path(
  "../data",
  eth_files
)
adm_level_labels <- stringr::str_extract(eth_files,"adm\\d")

ldf <- purrr::map(
  rlang::set_names(fp_eth_files, adm_level_labels),
  \(fp_tmp){
    arrow::read_parquet(
     fp_tmp
    )
  }
)

```

## At Country Level

### Plot of Monthly Forecast Means

Using a threshold, seasons can be defined as those months having mean rainfall above the quantile.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
sliderInput("season_thresh", "Threshold Quantile for Season:", 0.33, min = 0, max = 1)
adm0_monthly_means <- reactive({
  ldf$adm0 %>%
  mutate(month_name = factor(month.abb[valid_month], levels = month.abb)) %>%
  group_by(adm0_pcode, adm0_en, month_name) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  group_by(adm0_pcode, adm0_en) %>%
  mutate(season_thresh = quantile(mean_value, probs = input$season_thresh), 
         in_season = ifelse(mean_value >= season_thresh, "Yes", "No"))
})

renderPlot({
  ggplot(adm0_monthly_means(), aes(x=month_name, y=mean_value, group=1)) + 
    geom_line()+
    labs(title = "Average Forecasted Monthly Rainfall over Ethiopia", x = "Months", y = "Rainfall(mm)")+
    #geom_smooth(method = "loess", se = FALSE)+
    facet_wrap(~adm0_en)
})
```

### Table Showing Whether a Month is considered in season or not

```{r, echo=FALSE, message=FALSE, warning=FALSE}
DT::renderDT({
  DT::datatable(adm0_monthly_means() %>%
    select(month_name, in_season) %>% 
    pivot_wider(names_from = month_name, values_from = in_season), 
    filter = "top")
})
```

## At Regional Level

### Plot of Monthly Forecast Means

```{r, echo=FALSE, message=FALSE, warning=FALSE}
adm1_monthly_means <- reactive({
  ldf$adm1 %>%
  mutate(month_name = factor(month.abb[valid_month], levels = month.abb)) %>%
  group_by(adm1_pcode, adm1_en, month_name) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  group_by(adm1_pcode, adm1_en) %>%
  mutate(season_thresh = quantile(mean_value, probs = input$season_thresh), 
         in_season = ifelse(mean_value >= season_thresh, "Yes", "No"))
})

renderPlot({
  ggplot(adm1_monthly_means(), aes(x=month_name, y=mean_value, group=1)) + 
    geom_line()+
    labs(title = "Average Forecasted Monthly Rainfall over Ethiopia", x = "Months", y = "Rainfall(mm)")+
    facet_wrap(~adm1_en, scales = "free", ncol = 3)
})
```

### Table Showing Whether a Month is considered in season or not

```{r, echo=FALSE, message=FALSE, warning=FALSE}
DT::renderDT({
  DT::datatable(adm1_monthly_means() %>%
    select(month_name, in_season) %>% 
    pivot_wider(names_from = month_name, values_from = in_season),
    filter = "top")
})
```

## At Zonal Level

```{r, echo=FALSE, message=FALSE, warning=FALSE}
adm2_monthly_means <- reactive({
  ldf$adm2 %>%
  mutate(month_name = factor(month.abb[valid_month], levels = month.abb)) %>%
  group_by(adm1_pcode, adm1_en, adm2_pcode, adm2_en, month_name) %>%
  summarise(mean_value = mean(value, na.rm = TRUE)) %>%
  group_by(adm1_pcode, adm1_en, adm2_pcode, adm2_en) %>%
  mutate(season_thresh = quantile(mean_value, probs = input$season_thresh), 
         in_season = ifelse(mean_value >= season_thresh, "Yes", "No"))
})
```

### Table Showing Whether a Month is considered in season or not

```{r, echo=FALSE, message=FALSE, warning=FALSE}
DT::renderDT({
  DT::datatable(adm2_monthly_means() %>%
    select(month_name, in_season) %>% 
    pivot_wider(names_from = month_name, values_from = in_season),
    filter = "top")
})
```
